---
title: "INFO201 - PS5"
output: html_document
date: "2023-02-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**1 Load and check data (5pt)**

1. (1pt) For solving the problems, and answering the questions, create a new rmarkdown docu-ment with an appropriate title. 

2. (2pt) Load data. How many rows/columns do we have?

```{r}

library(dplyr)
library(ggplot2)

df <- read.delim("gapminder.csv")

nrow(df)
ncol(df)

```

3. (2pt) Print a small sample of data. Does it look OK?

```{r}

df %>% sample_n(5)

```


**2 Descriptive statistics (15pt)**

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r}

n_distinct(df$name)
n_distinct(df$iso2)
n_distinct(df$iso3)

```



2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
```{r}

df %>% group_by(iso2) %>% summarise(n = n_distinct(name)) %>% filter(n>1)
df %>% filter(iso2 =="" & name !="Namibia") %>% group_by(name) %>% summarise()

```
There is no iso-2 codes correspond to more than one name. In fact, there are two countries have blank iso-2 codes. One of them is Namibia, but that is because Namibia's iso-2 codes is 'NA'. There is one country has blank name and blank iso-2 code.


(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entitites are CHANISL and NLD CURACAO.
```{r}

df %>% filter(name == "" & name != "NA") %>% group_by(iso3) %>% summarise(names = unique(name))
                    
```
There are four iso3-codes correspond to blank country names, they are CHANISL, GBM, KOS, and NLD_CURACAO.


3. (2pt) What is the minimum and maximum year in these data?
```{r}

min(df$time,na.rm=TRUE)
max(df$time,na.rm=TRUE)

```
The minimum year is 1960 and the maximum year is 2019.



**3 CO2 emissions (30pt)**

Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}

df_missingco2 <- df %>% group_by(time) %>% summarise(n_missingc02 = sum(is.na(co2)),n_missingc02pc = sum(is.na(co2_PC)))

df_missingco2 %>% arrange(desc(n_missingc02)) 

```
It seems that 2017, 2018, and 2019 have the most missing data, they all have 217 missing data.


2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.
```{r}

co2_plot <- df %>% group_by(name) %>% summarise(total_co2 = sum(co2,na.rm=TRUE)) %>% filter(name == "United States of America" | name =="China" | name == "India" | name == "Japan" | name == "Australia")

ggplot(data = co2_plot, aes(y = total_co2, x = name ))+ geom_bar(stat="identity")

```
US has the most total CO2 emissions over time, China has the second most, then is Japan, India, and Australia. US and China's total CO2 emissions are significantly higher than the other three countries's.



3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). Make a similar plot of the same countries. What does this figure suggest?
```{r}

co2PC_plot <- df %>% group_by(name) %>% summarise(total_co2PC = sum(co2_PC,na.rm=TRUE)) %>% filter(name == "United States of America" | name =="China" | name == "India" | name == "Japan" | name == "Australia")

ggplot(data = co2PC_plot, aes(y = total_co2PC, x = name ))+ geom_bar(stat="identity")

```
US  has the most total CO2 emissions per capita, but this time Australia has the second most, then is Japan, China, and India. Since now CO2 is caculated by per capita, it make sense that country has larger population has less CO2 emissions per capita.


4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the same as continent). Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different size.
Hint: Americas 2016 should be 4.80.
```{r}

df_region <- df  %>% group_by(region, time) %>% summarise(avg_co2PC = mean(co2_PC, na.rm=TRUE)) 

df_region %>% arrange(desc(avg_co2PC))

df_region %>% filter(region == "Americas" & time == "2016") # check if match the hint

```
Ignoring the blank regions, Americas seems keep the CO2 emissions per capita in a high level, then is Europe and Asia.


5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita across continents in 1960 and 2016.
```{r}

df_region_1960_2016 <- df_region %>% filter(time == "1960" | time == "2016")
ggplot(data = df_region_1960_2016, aes(x = region, y = avg_co2PC, fill = factor(time), colour = factor(time))) + geom_bar(stat="identity",position = "dodge")

```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 percapita) in 2019 for each continent? (Assume region is continent).
```{r}

df_region %>% filter(time == "2019") 

```
I got all NA for filtering 2019, after remove the NAs, there is nothing left. I saw the conversation in Discord, it seems that I am not the only one with this problem. Otto said that it was fine to use the year 2016 instead as long as I explained. This chunk is to show that year 2019 does not work at all, and the next chunk will use year 2016.

```{r}

df_region2 <- df  %>% group_by(region, time) %>% summarise(avg_co2PC = mean(co2_PC, na.rm=TRUE))
df_region2 %>% filter(time == "2016") %>% arrange(desc(avg_co2PC)) #from largest to smallest
df_region2 %>% filter(time == "2016") %>% arrange(avg_co2PC) #from smallest to largest

```
The three largest CO2 emitters are Europe, Asia, and Americas (if ignore the blank region), and the three smallest CO2 emitters are Africa, Oceania, and Americas. Since there are only 5 region, Americas is one of the three largest and one of the three smallest.


**4 GDP per capita (50pt)**
Let’s look at GDP per capita (GDP_PC ).
1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r}

df %>% filter(time == "1960") %>% ggplot(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, col= region)) + geom_point()

```
European countries keep high GDP per capita and life expectancy on the same time, some American countries are on their way. Most African and Asian countries have low GDP per capita and low life expectancy. The plot suggests a positive relationship between GDP per capita and life expectancy.


2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r}

df %>% filter(time == "2019") %>% ggplot(aes(x = GDP_PC, y = lifeExpectancy, size = totalPopulation, col= region)) + geom_point()

```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

Over the 60 years, many Asian significantly improved their GDP per capita and life expectancy. American and European countries do not have significant changes except their dots become smaller in size but increase in total number. Some African countries improved their popualtion as well as the life expectancy. All countries have improvements.


4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of different size.
```{r}

df %>% filter(time == "1960" | time == "2019") %>% group_by(region, time) %>% summarise(avgGDP_PC = mean(GDP_PC, na.rm = TRUE), avglifeExpectancy = mean(lifeExpectancy, na.rm = TRUE))

```

Yes, the results fit with what I see on the figures. All continent improve their average GDP per capita and life expectancy, especially Asia.


5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.
```{r}

df %>% group_by(region, time) %>% summarise(avgLE = mean(lifeExpectancy, na.rm = TRUE)) %>% mutate(prevLE = lag(avgLE), growth = avgLE - prevLE) 


```
Africa's LE grew quickly initially, but the speed slowed down since 1986 and some times even have negative growth. Similarly, Americas has a high growth in the first decade and then slowed down. Asia has the highest growth in the first decade, but then the growth gradually go down to the normal level as other continents. Europe's growth is unstable and have more negative growth than Africa. Oceania's growth keeps between 0.2 - 0.3.


6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}

df %>% filter(time == "1960" | time == "2019") %>% group_by(time) %>% summarise(average_GDP_PC = mean(GDP_PC, na.rm = TRUE)) %>% ggplot(aes(x = average_GDP_PC, fill = factor(time), col = factor(time) )) + geom_histogram()

```
Here I assume that the question is asking for the year's average GDP per capita for 1960 and 2019.



7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)

Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r}

df %>%  filter(time == "1960" & name != "") %>% mutate(rank = rank(-lifeExpectancy) ) %>% filter (name == "United States of America")

df %>%  filter(time == "2019"& name != "") %>% mutate(rank = rank(-lifeExpectancy) ) %>% filter (name == "United States of America")

```


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019–what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960.
```{r}

df_1960rank <- df %>%  filter(time == "1960" & name != "" & !is.na(lifeExpectancy)) %>% mutate(rank = rank(-lifeExpectancy) ) 
df_1960rank[df_1960rank$name == "United States of America", "rank"]/ n_distinct(df_1960rank$name)

df_2019rank <- df %>%  filter(time == "2019" & name != "" & !is.na(lifeExpectancy)) %>% mutate(rank = rank(-lifeExpectancy) ) 
df_2019rank[df_2019rank$name == "United States of America", "rank"]/ n_distinct(df_2019rank$name)

```


**Finally** tell us how many hours did you spend on this PS.

I spend 6-7 hours on this PS. The hardest part is to understand what the question is asking...
